---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "lastName, firstName"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk. Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed. You may use either base R functions or ggplot2 for the visualizations.

------------------------------------------------------------------------

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}
# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

------------------------------------------------------------------------

### Test Items starts from here - There are 6 sections - Total 50 points

***\#\#\#\#\# Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata. Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.

```{r Part_1a}
library(tidyverse)
summary(mydata)
table(select(mydata, CLASS, RINGS))

```

**Question (1 point): Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (Based on the dable above I would say that class is based on rings. A5 looks to contain alot of outliers in class 5 and classes 2-4 contain most of the meat of the data in this context. Based on just this frequency table I probably would do my analysis based on each class type considering the size of the A5 RINGS distribution and present it that way. I would aggregate if the clients really insisted)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)). Lastly, present a barplot of these data; ignoring the marginal totals.

```{r Part_1b}
df <- table(select(mydata, SEX, CLASS))
addmargins(df)
barplot(df, main='Sex & Class Barchart', xlab='CLASS', legend = TRUE)
```

**Essay Question (2 points): Discuss the sex distribution of abalones. What stands out about the distribution of abalones by CLASS?**

***Answer: (First thing that catches my eye is that we do not have even numbers of Sex and class which could skew the data. I wonder if this is true of the overall population of this species or if we are introducing bias in this manner. I'de want to know how they were caught and how the data was measured; also I want to know the geographical location and timeframes that they were caught. I dont know squat about marine life but for all we know they could have been catching in an area or timeframe where certian sexes are less or more common due to something like the mating practices of the species.I would call an expert on marine biology to ensure the data collection method didn't introduce bias. If there was no bias and this was representative of the population I would say that there is a significantly higher distribution of the SEX type I in the lower classes with even M & F throughout.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87).

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment.

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata), 200, replace=FALSE),] 
plot(work[, 2:6])
```

------------------------------------------------------------------------

***\#\#\#\#\# Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = WHOLE, y = VOLUME)) + geom_point(aes(color = CLASS),size = 3) + theme(plot.title = element_text(color = "blue", size = 12, face = "bold", hjust = 0.5))+ labs(title="WHOLE versus VOLUME")
```

(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE. Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
max_ratio <- max(mydata$SHUCK / mydata$WHOLE)
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + geom_point(aes(color = CLASS),size = 3) + theme(plot.title = element_text(color = "blue", size = 12, face = "bold", hjust = 0.5))+ labs(title="WHOLE versus SHUCK")+geom_abline(slope=max_ratio, intercept=0)


```

**Essay Question (2 points): How does the variability in this plot differ from the plot in (a)? Compare the two displays. Keep in mind that SHUCK is a part of WHOLE. Consider the location of the different age classes.**

***Answer: (The linear distrubition is tighter for A5, its not great but its better than in the first graph. This makes sense because as you said, shuck is a part of whole)***

------------------------------------------------------------------------

***\#\#\# Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.

```{r Part_3a}
head(mydata)
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX=='M'], col = "blue", main='M', xlab='Ratio')
hist(mydata$RATIO[mydata$SEX=='F'], col = "red", main='F', xlab='Ratio')
hist(mydata$RATIO[mydata$SEX=='I'], col = "green", main='I', xlab='Ratio')
boxplot(mydata$RATIO[mydata$SEX=='M'], col = "blue", main='M')
boxplot(mydata$RATIO[mydata$SEX=='F'], col = "red", main='F')
boxplot(mydata$RATIO[mydata$SEX=='I'], col = "green", main='I')
qqnorm(mydata$RATIO[mydata$SEX=='M'], col = "blue", main='M')
qqline(mydata$RATIO[mydata$SEX=='M'], col = "black")
qqnorm(mydata$RATIO[mydata$SEX=='F'], col = "red", main='F')
qqline(mydata$RATIO[mydata$SEX=='F'], col = "black")
qqnorm(mydata$RATIO[mydata$SEX=='I'], col = "green", main='I')
qqline(mydata$RATIO[mydata$SEX=='I'], col = "black")
```

**Essay Question (2 points): Compare the displays. How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (The qq plots indicate a right skewed population which I was worried about earlier; this is especially true for the SEX I as it has the fattest tail. SEX F also has some undesirable outliers per the box plot.)***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
# {boxplot.stats(mydata$RATIO[mydata$SEX=='M'])
# {boxplot.stats(mydata$RATIO[mydata$SEX=='F'])
# {boxplot.stats(mydata$RATIO[mydata$SEX=='I'])

kable(subset(mydata, mydata$RATIO %in% boxplot(mydata$RATIO, plot=FALSE)$out))

```

**Essay Question (2 points): What are your observations regarding the results in (3)(b)?**

***Answer: (I would say there is a large amount of outliers within the infant sex.)***

------------------------------------------------------------------------

***\#\#\# Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots: VOLUME and WHOLE versus RINGS. Present these four figures in one graphic: the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.

```{r Part_4a}

par(mfrow = c(2,5))
A1<-mydata[mydata$CLASS=='A1',]
A2<-mydata[mydata$CLASS=='A2',]
A3<-mydata[mydata$CLASS=='A3',]
A4<-mydata[mydata$CLASS=='A4',]
A5<-mydata[mydata$CLASS=='A5',]
boxplot(A1$VOLUME, A1$WHOLE, col = "blue", main='A1', names = c("V", "W"))
boxplot(A2$VOLUME, A2$WHOLE, col = "red", main='A2', names = c("V", "W"))
boxplot(A3$VOLUME, A3$WHOLE, col = "green", main='A3', names = c("V", "W"))
boxplot(A4$VOLUME, A4$WHOLE, col = "yellow", main='A4', names = c("V", "W"))
boxplot(A5$VOLUME, A5$WHOLE, col = "purple", main='A5', names = c("V", "W"))
plot(mydata$VOLUME, mydata$RINGS)
plot(mydata$WHOLE, mydata$RINGS)
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age? Explain.**

***Answer: (I believe that CLASSES are a good predicator of age in the data set. I am not really seeing any use to the graphs above.)***

------------------------------------------------------------------------

***\#\#\# Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose. You do not need to be concerned with the number of digits presented.

```{r Part_5a}
datarat <- aggregate(RATIO ~ SEX + CLASS, data = mydata, mean)
datavol <- aggregate(VOLUME ~ SEX + CLASS, data = mydata, mean) 
datashuck <- aggregate(SHUCK ~ SEX + CLASS, data = mydata, mean)

mat_data <-matrix(data=datarat[,3], nrow = 3, byrow = FALSE)
colnames(mat_data) <- c("A1","A2","A3","A4","A5")
rownames(mat_data) <- c("Ratio_F", "Ratio_I", "Ratio_M")

vol_data <-matrix(data=datavol[,3], nrow = 3, byrow = FALSE)
colnames(vol_data) <- c("A1","A2","A3","A4","A5")
rownames(vol_data) <- c("Volume_F", "Volume_I", "Volume_M")

shuc_data <-matrix(data=datashuck[,3], nrow = 3, byrow = FALSE)
colnames(shuc_data) <- c("A1","A2","A3","A4","A5")
rownames(shuc_data) <- c("Shuck_F", "Schuck_I", "Schuck_M")
mat_data
vol_data
shuc_data
```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
par(mfrow=c(3,1))
interaction.plot(response = mydata$RATIO, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, trace.label = "sex", xlab= "class", ylab = "ratio", col= 2:5)

interaction.plot(response = mydata$VOLUME, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, trace.label = "sex", xlab= "class", ylab = "volume", col= 2:5)

interaction.plot(response = mydata$SHUCK, trace.factor = mydata$SEX, x.factor = mydata$CLASS, fun = mean, trace.label = "sex", xlab= "class", ylab = "shuck", col= 2:5)

```

**Essay Question (2 points): What questions do these plots raise? Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS \< 16. Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE. If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2, 2))
a <- subset(mydata, mydata$RINGS < 16)
boxplot(a$VOLUME[a$SEX == "I"] ~ a$RINGS[a$SEX == "I"], ylim=c(0,1100), main= "I Volume - Rings", col= "Blue", Xlab = "Rings")
boxplot(a$VOLUME[a$SEX != "I"] ~ a$RINGS[a$SEX != "I"], ylim=c(0,1100), main= "M&F Volume-Rings", col= "Green", Xlab = "Rings")
boxplot(a$WHOLE[a$SEX == "I"] ~ a$RINGS[a$SEX == "I"], ylim=c(0,400), main= "I Whole Weight-Rings", col= "Blue", Xlab = "Rings")
boxplot(a$WHOLE[a$SEX != "I"] ~ a$RINGS[a$SEX != "I"], ylim=c(0,400), main= "M&F Whole Weight-Rings", col= "Green", Xlab = "Rings")   

```

**Essay Question (2 points): What do these displays suggest about abalone growth? Also, compare the infant and adult displays. What differences stand out?**

***Answer: (I would say that there is a correlation between volume and weight. At around age 12 it begins to fluctuate for infants too.)***

------------------------------------------------------------------------

***\#\#\# Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***

**Conclusions**

**Essay Question 1) (5 points) Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (It is clear that there is bias within the data set. There are an unacceptable number of outliers concerning the SEX of the animals. I believe that the data was not gathered well and there needs to be stronger control around the process.)***

**Essay Question 2) (3 points) Do not refer to the abalone data or study. If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would take alot of concern as to how the data was gathered. I would want to know the time, place, circumstance and any other bias that the data collection might introduce.)***

**Essay Question 3) (3 points) Do not refer to the abalone data or study. What do you see as difficulties analyzing data derived from observational studies? Can causality be determined? What might be learned from such studies?**

***Answer: (Absolutely, I believe that is the number one proble of past methods of data collection rather than IOT devices. You can gage a fair amount from data gathered this way but I think that more advanced statistical analysis, specifically machine learning applications is quite useless, at least that has been my experience. Humans suck at logging data, myself included)***
